Computational design of RNA-based oscillatory circuits

### 2015-09-13 MScWriteup.tex
still rewriting last section

\begin{align}
\frac{\mathrm{d}\mathbf{y}}{\mathrm{d}t} = \mathbf{f}(\mathbf{y},\boldsymbol{\theta},t),
\end{align}
where $\mathbf{y}$ is our state vector, $\boldsymbol{\theta}$ is a vector of model parameters, and $t$ is time. The model may be integrated numerically, giving a prediction $\mathbf{y}(t,\boldsymbol{\theta})$. The least squares error between the model prediction and an experimental time series is defined as
\begin{align}
J(\boldsymbol{\theta}) = \sum_{i =1}^{n} (\mathbf{y}_{\mathrm{exp}}(t_{i}) - \mathbf{y}(t_{i},\boldsymbol{\theta}))^2,
\end{align}
where the experimental time series, $\mathbf{y}_{\mathrm{exp}}(t_{i})$ is recorded at timepoints $t_{i}$, $i = 1 \hdots n$. This error function defines a landscape in $\boldsymbol{\theta}$ space, and we seek to minimise it by varying $\boldsymbol{\theta}$. In our case, we do not have experimental data on the full state vector, but only one component of it - the observed fluorescence, $z(t)$. In addition, rather than a single experimental run, we have many, corresponding to a time series from each cell. We incorporate this by fitting to the experimental mean of the data, and only minimising over the observed component. Our minimisation problem is thus
\todo[inline]{explain why not fit each curve individually.}
\begin{align}
\min_{\boldsymbol{\theta}} \sum_{i =1}^{n} (z_{\mathrm{exp, mean}}(t_{i}) - z(t_{i},\boldsymbol{\theta}))^2.
\label{eq:min}
\end{align}
The next step is performing the minimisation. In general, the landscape defined by the error function may be rugged and contain many local minima, which a local optimisation algorithm will may get stuck in \footnote{this is true even if ODE model is linear in its parameters, as is almost the case for us - though the ODE model is linear, the resulting solutions are in general not. A counterexample is the harmonic oscillator.}. To try and surmount this problem, \cite{Algorithms2003} suggests the use of a global optimisation algorithm, and in particular recommends several Evolutionary Algorithms, of which we choose one, the CMA-ES \cite{Hansen2006,Hansen2011}. \footnote{CMA-ES stands for covariance matrix evolutionary strategy. This algorithm generates a population of test points according to a multivariate gaussian distribution. It then ranks the points based on their fitness function scores, and  considers some number of the highest ranked points as a sample from a new multivariate gaussian. Using this sample, it then estimates a new mean and covariance matrix for this gaussian and the process is iterated.}
\todo[inline]{why use CMA-ES? Some refs}

@@

Figs. \ref{InitialResults_13_9_correlation}, \ref{InitialResults_14_7_correlation} show correlation matrices between parameter sets, computed by viewing each of the estimated parameter sets as a sample from a multivariate random variable. An element of the correlation matrix $R_{ij}$ is then the Pearson correlation coefficient between the $i^\mathrm{th}$ and $j^\mathrm{th}$ random variable - or, in our case, the $i^\mathrm{th}$ and $j^\mathrm{th}$ parameters. These correlation matrices indicate that there are spaces of parameters within which the fitness function remains approximately constant - for example, in both datasets there exists of positive correlation between values of $\mu$ and $\beta$. Referring to \eqref{eq:p}, this makes heuristic sense - the two parameters may have similar effects on model predictions, and may be able to co-vary in such a way as to leave the model prediction unchanged.
The results suggest a fitness landscape relatively flat to perturbations in certain combinations of parameters, and indicate we may have difficulty obtaining unique estimates of the model parameters.

@@

 There are two main reasons why a parameter may not be estimable \cite{Mclean2012,Yao2003, Beck}: Model predictions may be insensitive to the value of a particular parameter, or the effects of varying one parameter on model predictions may be highly correlated with the effects of varying several others.

@@

Problems may also arise for more practical reasons (\textit{practical identifiability} \cite{Mclean2012}). For example, it is possible that in the experimental regime we operate in, parameter effects may be weak, or highly correlated, but in other regimes this is not true. \footnote{\cite{Beck} gives an example in which a parameter only affects model predictions after several hours, though others will affect it at all timescales - in this example, if we only took data for a few minutes, the parameter would be inestimable, but in principle it is not.} In this case, parameter estimates may be improved by taking data in more varied experimental conditions, and attempting to observe as many components of the model output as possible.

@@

@@

where $F$ is a phenomenological scaling factor. We now have three free parameters, $F,\mu,\Theta$, which we may estimate as before. Results are shown in Fig.\ref{SimplifiedModelResults}. In contrast to the results of Fig.\ref{InitialResults}, we see relatively tightly estimated parameters, with no estimated parameter values near the initial bounding box. Further, this simplified model achieves error values as low as the more complex model, and predictions from both are similar. In fact, the simplified algorithm is more consistently able to find the minimal function value identified in Fig. \ref{Initial fit Error}, possibly because the low dimension of the new model makes parameter space easier to search. We also note positive correlation between $F$ and $\Theta$, suggesting that both play similar roles in determining the fit. This may be explained as follows: neglecting the ClpX degradation term, \eqref{eq:p_simplified} is a simple exponential decay to a fixed value. Here, $F$ will determine the value the $p$ will tend to, and $\mu$ will determine its rate constant. $F$ is then a linear scaling factor, just as $\Theta$ is - both scale the value the exponential will tend to.

@@

>>>>

\begin{align}
\frac{\mathrm{d}\mathbf{y}}{\mathrm{d}t} = \mathbf{f}(\mathbf{y},\boldsymbol{\theta},t),
\end{align}
where $\mathbf{y}$ is our state vector, $\boldsymbol{\theta}$ is a vector of model parameters, and $t$ is time. The model may be integrated numerically, giving a prediction $\mathbf{y}(t,\boldsymbol{\theta})$. The least squares error between the model prediction and an experimental time series is defined as
\begin{align}
J(\boldsymbol{\theta}) = \sum_{i =1}^{n} (\mathbf{y}_{\mathrm{exp}}(t_{i}) - \mathbf{y}(t_{i},\boldsymbol{\theta}))^2,
\end{align}
where the experimental time series, $\mathbf{y}_{\mathrm{exp}}(t_{i})$ is recorded at timepoints $t_{i}$, $i = 1 \hdots n$. This error function defines a landscape in $\boldsymbol{\theta}$ space, and we seek to minimise it by varying $\boldsymbol{\theta}$. In our case, we do not have experimental data on the full state vector, but only one component of it - the observed fluorescence, $z(t)$. In addition, rather than a single experimental run, we have many, corresponding to a time series from each cell. We incorporate this by fitting to the experimental mean of the data, and only minimising over the observed component. Our minimisation problem is thus
\todo[inline]{explain why not fit each curve individually.}
\begin{align}
\min_{\boldsymbol{\theta}} \sum_{i =1}^{n} (z_{\mathrm{exp, mean}}(t_{i}) - z(t_{i},\boldsymbol{\theta}))^2.
\label{eq:min}
\end{align}
The next step is performing the minimisation. In general, the landscape defined by the error function may be rugged and contain many local minima, which a local optimisation algorithm will may get stuck in. To try and surmount this problem, \cite{Algorithms2003} suggests the use of a global optimisation algorithm, and in particular recommends several Evolutionary Algorithms, of which we choose one, the CMA-ES \cite{Hansen2006,Hansen2011}. 
\todo[inline]{why use CMA-ES? Some refs}

@@

Figs. \ref{InitialResults_13_9_correlation}, \ref{InitialResults_14_7_correlation} show correlation matrices between parameter sets\footnote{These are computed by viewing each of the estimated parameter sets as a sample from a multivariate random variable. An element of the correlation matrix $R_{ij}$ is then the Pearson correlation coefficient between the $i^\mathrm{th}$ and $j^\mathrm{th}$ random variable - or, in our case, the $i^\mathrm{th}$ and $j^\mathrm{th}$ parameters}. These correlation matrices indicate that there are spaces of parameters within which the fitness function remains approximately constant - for example, in both datasets there exists of positive correlation between values of $\mu$ and $\beta$. Referring to \eqref{eq:p}, this makes heuristic sense - the two parameters may have similar effects on model predictions, and may be able to co-vary in such a way as to leave the model prediction unchanged.
The results suggest a fitness landscape relatively flat to perturbations in certain combinations of parameters, and indicate we may have difficulty obtaining unique estimates of the model parameters.

@@

 There are two main reasons why a parameter may not be estimable \cite{Mclean2012,Yao2003, Beck,Jimenez-Hornero2008,Grewal1976}: Model predictions may be insensitive to the value of a particular parameter, or the effects of varying one parameter on model predictions may be highly correlated with the effects of varying several others.

@@

Problems may also arise for more practical reasons (\textit{practical identifiability} \cite{Mclean2012}). For example, it is possible that in the experimental regime we operate in, parameter effects may be weak, or highly correlated, but in other regimes this is not true. In this case, parameter estimates may be improved by taking data in more varied experimental conditions, or attempting to increase the number of model outputs observed - in our case, observing more of the state vector than simply $z$.

@@

\subsection{A simplified model}

@@

where $F$ is a phenomenological scaling factor. We now have three free parameters, $F,\mu,\Theta$, which we may estimate as before. Results are shown in Fig.\ref{SimplifiedModelResults}, for the \textit{13\_9} dataset only. In contrast to the results of Fig. \ref{InitialResults}, we see relatively tightly estimated parameters, with no estimated parameter values near the initial bounding box \footnote{Initial bounding box values given in appendix \ref{Simplified model initial bounding box}.}. Further, this simplified model achieves error values as low as the more complex model, and predictions from both are similar. In fact, the simplified algorithm is more consistently able to find the minimal function value identified in Fig. \ref{Initial fit Error}, possibly because the low dimension of the new model makes parameter space easier to search. We also note positive correlation between $F$ and $\Theta$, suggesting that both play similar roles in determining the fit. Fig. \ref{Sensitivity_scaled_simplified} shows the normalised sensitivity curves using this simplified model for the \textit{13\_9} dataset, and also suggests $F$ and $\Theta$ play similar, though not identical, roles.

  This may be explained as follows: neglecting the ClpX degradation term, \eqref{eq:p_simplified} is a simple exponential decay to a fixed value. Here, $F$ will determine the value the $p$ will tend to, and $\mu$ will determine its rate constant. $F$ is then a linear scaling factor, just as $\Theta$ is - both scale the value the exponential will tend to.

\begin{figure}[h]
\centering
\includegraphics[scale = 0.23,clip = true, trim = 90 0 60 0]{Sensitivity_scaled_simplified}
\caption{Using the simplified model \eqref{eq:p_simplified} - \eqref{eq:z_simplified}, normalised sensitivity coefficients $S_{ij}$ evaluated about a set of estimated parameters from the \textit{13\_9} dataset, for a single oscillation.}
\label{Sensitivity_scaled_simplified}
\end{figure}

@@

\section{Simplified model initial bounding box}
\label{Simplified model initial bounding box}
\setcounter{figure}{0}

\begin{table}[h!]
\renewcommand{\arraystretch}{1.3}
\caption{Initial bounds on parameters to be estimated in the simplified model, \eqref{eq:p_simplified} - \eqref{eq:z_simplified}.}
\centering
\begin{tabular}{| l | l | p{0.3\linewidth} | p{0.1\linewidth} | p{0.1\linewidth} |}
\hline \textbf{Parameter} &  \textbf{Units} & \textbf{Definition}  & \textbf{Initial Bounds}  \\
\hline ${\mu}$ &  /min & Dilution rate  & 0.001-0.1\\
\hline ${F}$ &  nM /min & Phenomenological forcing  & 10-1500\\
\hline ${\Theta}$  &   nM/AFU & Ratio between GFP concentration and observed fluoresence  & 100 - 10000  \\
\hline
\end{tabular}
\end{table}
